Dynamical dark energy constraints via clustering shells of SDSS-III BOSS galaxies

### 2017-07-13 Draft/Draft0.6.tex

\vspace{-3.5mm}
\begin{abstract}
%\sffamily
We use the small scale anisotropic clustering signal of 1,133,326 galaxies from the Sloan Digital Sky Survey Data Release 
12, spanning the redshift range $0.15<z<0.69$ to place constraints on the expansion history of the Universe. 
Concentrating on the Chevallier-Polarski-Linder parametrization of dark energy, $w=w_0+w_a\frac{z}{1+z}$,
while improving upon earlier works \citep{Li2016}, 
we present a novel use of the Alcock-Paczynski (AP) test via the small scales clustering difference between redshift bins. 
This differential measure is shown to be a robust statistical tool that mitigates many systematic effects, 
allowing for competitive and unbiased constrains to be placed on cosmological parameters.
We find that our results, in combination with the CMB, BAO, SNIa and $H_0$ from Cepheid data yield, 
$\Omega_m = 0.301 \pm 0.008,\ w_0 = -1.042 \pm 0.067,\ w_a = -0.07 \pm 0.29$ (68.3\% CL). 
Adding our new AP measurements to the aforementioned results reduced the $w_0-w_a$ 1-$\sigma$ contour area by 50\%.
The AP method can be combined with the standard BAO method, 
increasing the science output of ongoing and future galaxy surveys by a factor of 2.
\end{abstract}

@@

Previously, we proposed to use the  {\it redshift dependence} of the AP distortion\cite{Li2014} as a way of mitigating the RSD effect. The clustering anisotropies produced by RSD are, although large, close to uniform in magnitude over a wide range in redshift.  However, if cosmological parameters are incorrectly chosen resulting in an AP effect, 
the anisotropy in the clustering signal has a clear redshift dependence. 
In a later work\cite{Li2015}, we developed an AP methodology 
that utilises the redshift dependence of the galaxy 2-point correlation function (2pCF), 
measured as a function of angle between the galaxy pair and line-of-sight (LoS).
As an illustration, Fig.\ref{fig_xy} shows how the shape distortion of objects (anisotropy of clustering signal) varies with distance (or redshift) when incorrect values of  $w_0$ or $w_a$ are used to infer distances.

In earlier work\citep{Li2016} we applied this AP method to galaxies from the 3rd incarnation of the Sloan Digital Sky Survey (SDSS-III). Combining their method with measurements of the Cosmic Microwave Background (CMB), type Ia supernovae (SNIa), 
baryon acoustic oscillations (BAO), and $H_0$,
they obtained very tight constraints of $ \Omega_m = 0.301 \pm 0.006,\ w=-1.054 \pm 0.025$.
In reducing the RSD effect, 
they were able to use galaxy clustering on scales down to 6 $h^{-1}$Mpc,
which is a major advance in extracting cosmological information 
on small scales where galaxy clustering is strong and there are many independent structures.

In this paper, we continue to develop our previous methodology and proceed to infer the constraints on dynamical dark energy. 
We will present a slightly improved methodology to that of our previous work\citep{Li2016}, 
allowing for faster likelihood estimation and thus the exploration of larger, higher dimensional parameter spaces. 
The methodology we will present here, can be applied to any model of dynamical dark energy, 
or indeed any appropriately chosen parametric or non-parametric decomposition of the cosmic expansion history. 
However, as a first step, in this paper we will focus on the widely used Chevallier-Polarski-Linder (CPL) parametrization
$w(z) = w_0 + w_a \frac{z}{1+z}$.

The Planck team has released the {\texttt {COSMOMC}} \citep{LB2002} outputs of four MCMC ``chains'' in the CPL model, 
using a combination of four datasets:
the full-mission Planck observations of CMB temperature and polarization anisotropies \cite{Planck2015},
the BAO distance priors measured from SDSS DR11 \citep{Anderson2013}, 6dFGS \citep{6dFGS} and SDSS MGS \citep{MGS},
the ``JLA'' SNIa sample \citep{JLA},
and the Hubble Space Telescope measurement of $H_0=70.6\pm3.3$ \cite{Riess2011,E14H0}.
These MCMC chains contain the CMB+BAO+SNIa+$H_0$ likelihood computed for $\sim$37,000 sets of cosmological parameters,
yielding constraints of $ \Omega_m = 0.309 \pm 0.010, w_0 = -0.938 \pm 0.109, w_a = -0.38 \pm 0.41$.
%\begin{equation}
 %\Omega_m = 0.309 \pm 0.010, w_0 = -0.938 \pm 0.109, w_a = -0.38 \pm 0.41.
%end{equation}
We then compute the likelihood of the LOWZ+CMASS AP measurements for each of the parameter combinations provided in the Planck chain. 
To do this, we splited the 1,133,326 galaxies from the SDSS DR12 into six redshift bins,
measured the 2pCF in these redshift bins, 
subtract the systematic effects based on estimations from mocks made from Horizon Run 4 simulations \citep{HR4},
and use a $\chi^2$ function to assess the redshift dependence
(see Methods for further explanations).
After adding the sample log-likelihoods with ours, 
while also multiply the sample weights by our likelihoods, 
we derive the CMB+BAO+SNIa+$H_0$+AP constraints on CPL parameters, $\Omega_m = 0.301 \pm 0.008, w_0 = -1.042 \pm 0.067, w_a = -0.07 \pm 0.29$.
%\begin{equation}
%\Omega_m = 0.301 \pm 0.008, w_0 = -1.042 \pm 0.067, w_a = -0.07 \pm 0.29.
%\end{equation}
While the central values of parameters do not shift appreciably,
the error bars of $w_0$ and $w_a$ are reduced by 38.4\% and 29.7\%, respectively.% $$-40\%$ by adding our method into the analysis.

@@

The presented method of clustering shells can be applied to many spectroscopic galaxy surveys. 
The method is complimentary to the BAO method and the two methods can be combined together, 
increasing (by roughly twice) the science output of ongoing and future surveys, with little to no additional cost.
We expect this approach will play an important role in helping us unveiling the mystery of dark energy.

@@

@@

X.L. computed the approximated 2pCFs in the many cosmologies, did the statistical analysis and computed the likelihoods, 
derived the final cosmological constraints, and conducted the robustness tests. 
C.S. wrote the code for computing the 2pCFs, and contributed in the likelihood methodology and the robustness tests. 
C.P. contributed in the concept of using the redshift dependence, the LSS statistical, the likelihood methodology,
the horizon run simulations, the systematic corrections, and the robustness tests.
Y.W. and G.Z. contributed in the computing of approximated 2pCFs in non-fiducial cosmologies and the robustness tests.
H.P. contributed in the 2pCF measurements, the likelihood computation, and the systematic correction.
A.S. contributed in the fast computation of likelihoods in non-fiducial cosmologies and the cosmological interpretation of the results.
J.K. and S.H. performed the Horizon Run 4 simulations and created the mock galaxy samples.
All authors commented on the manuscript.

@@

In detail, the number of pairs are only counted in a fiducial cosmology
-- which is taken as the $\Omega_m=0.26$ $\Lambda$CDM --
and ``translated'' to the measurements in a ``target'' cosmology using the relations, 
\begin{eqnarray}
 s_{\rm target} = s_{\rm fiducial} \sqrt{\alpha_{\parallel}^2 \mu_{\rm fiducial}+\alpha_{\bot}(1-\mu_{\rm fiducial}^2)}, \\
 \mu_{\rm target} = \mu_{\rm fiducial} \frac{\alpha_\parallel}
 {\sqrt{\alpha_{\parallel}^2 \mu_{\rm fiducial} +\alpha_{\bot}(1-\mu_{\rm fiducial}^2)}}
\end{eqnarray}
where $\alpha_{\bot}\equiv D_{A,\rm target}/D_{A,\rm fiducial}$,
$\alpha_{\parallel}\equiv H_{\rm fiducial}/H_{\rm target}$.
and values of $D_A$ and $H$ are computed in the effective redshifts of the six redshift bins.
In the fiducial cosmology
we measure $\xi(s,\mu)$ in an ultra high resolution of
$\Delta s = 0.2 {\rm Mpc/h}$, $\Delta \mu = 1/600$,
%Let us call this ``approximate 2pCF'' method.
%to obtain the accurate number counts 
and these small ``pixels'' are grouped to infer 
the number counts in other cosmologies, 
in large pixels of $\Delta s = 1 {\rm Mpc/h}$, $\Delta \mu = 1/120$.
%in the fiducial cosmology and group the bins in the target cosmology.
In the case when one small pixel belongs to more than one larger pixel,
a correction is applied by computing the fraction of the overlapping area.
We found when using the approximate method the computed 
$\hat\xi_{\Delta s}(\mu)$ in a non-fiducial cosmology suffers from
an error of $\approx0.5\%$, which is 10 times smaller than 
the statistical noise in the 2pCF.

>>>>

\vspace{-3.5mm}
\begin{abstract}
%\sffamily
We use the small scale anisotropic clustering signal of 1,133,326 galaxies from the Sloan Digital Sky Survey Data Release 
12, spanning the redshift range $0.15<z<0.69$ to place constraints on the expansion history of the Universe. 
Concentrating on the Chevallier-Polarski-Linder parametrization of dark energy, $w=w_0+w_a\frac{z}{1+z}$,
while improving upon earlier works \citep{Li2016}, 
we present a novel use of the Alcock-Paczynski (AP) test via the small scales clustering difference between redshift bins. 
This differential measure is shown to be a robust statistical tool that mitigates many systematic effects, 
allowing for competitive and unbiased constrains to be placed on cosmological parameters.
We find that our results, in combination with the CMB, BAO, SNIa and $H_0$ from Cepheid data yield, 
$\Omega_m = 0.301 \pm 0.008,\ w_0 = -1.042 \pm 0.067,\ w_a = -0.07 \pm 0.29$ (68.3\% CL). 
Adding our new AP measurements to the aforementioned results reduced the $w_0-w_a$ 2-$\sigma$ contour area by 50\%.
The AP method can be combined with the standard BAO method, 
greatly increasing the power of ongoing and future galaxy surveys in probing dark energy.
\end{abstract}

@@

Previously, we proposed to use the  {\it redshift dependence} of the AP distortion\cite{Li2014} as a way of mitigating the RSD effect. The clustering anisotropies produced by RSD are, although large, close to uniform in magnitude over a wide range in redshift.  However, if cosmological parameters are incorrectly chosen resulting in an AP effect, 
the anisotropy in the clustering signal has a clear redshift dependence
(as an illustration, Fig.\ref{fig_xy} shows how the shape distortion of objects varies with distance (i.e., redshift) 
when incorrect cosmologies are used to infer distances).
In a later work\cite{Li2015}, we developed an AP methodology 
that utilises the redshift dependence of the galaxy 2-point correlation function (2pCF), 
measured as a function of angle between the galaxy pair and line-of-sight (LoS).

In an earlier work\citep{Li2016} we applied this AP method to galaxies from the 3rd incarnation of the Sloan Digital Sky Survey (SDSS-III). 
Combining the method with measurements of the Cosmic Microwave Background (CMB), type Ia supernovae (SNIa), 
baryon acoustic oscillations (BAO), and $H_0$,
we obtained very tight constraints of $ \Omega_m = 0.301 \pm 0.006,\ w=-1.054 \pm 0.025$.
In reducing the RSD effect, 
we were able to use galaxy clustering on scales down to 6 $h^{-1}$Mpc,
which is a major advance in extracting cosmological information 
on small scales where galaxy clustering is strong and there are many independent structures.

In this paper, we continue to develop our previous methodology and proceed to infer the constraints on dynamical dark energy. 
We will present a slightly improved methodology to that of our previous work\citep{Li2016}, 
allowing for faster likelihood estimation and thus the exploration of larger, higher dimensional parameter spaces
(see Methods for details). 
The methodology we will present here, can be applied to any model of dynamical dark energy, 
or indeed any appropriately chosen parametric or non-parametric decomposition of the cosmic expansion history. 
However, as a first step, in this paper we will focus on the widely used Chevallier-Polarski-Linder (CPL) parametrization\cite{CPL_CP,CPL_L},
\begin{equation}
w(z) = w_0 + w_a (1-a) = w_0 + w_a \frac{z}{1+z}.
\end{equation}
This parametrization characterizes dark energy equation-of-state (EoS) by two free parameters.
$w_0$ determines the present-day value, while parameter $w_a$ characterizes the first-order derivative of $w$ with respect to $a$. 
The CPL parametrization has many obvious advantages, for instance, a manageable parameter space, 
the bounded behavior at high redshift, 
and the ability to accurately reconstruct many dark energy theories\cite{CPL_L}. 
So it was widely used in the literature to reconstruct dark energy from observations.

%The CPL parametrization has been widely used to study the dynamical properties of dark energy in literature.
The Planck team has released the {\texttt {COSMOMC}} \citep{LB2002} outputs of four MCMC ``chains'' in the CPL model, 
using a combination of four datasets:
the full-mission Planck observations of CMB temperature and polarization anisotropies \cite{Planck2015},
the BAO distance priors measured from SDSS DR11 \citep{Anderson2013}, 6dFGS \citep{6dFGS} and SDSS MGS \citep{MGS},
the ``JLA'' SNIa sample \citep{JLA},
and the Hubble Space Telescope measurement of $H_0=70.6\pm3.3$ \cite{Riess2011,E14H0}.
These MCMC chains contain the CMB+BAO+SNIa+$H_0$ likelihood computed for $\sim$37,000 sets of cosmological parameters,
yielding constraints of %$ \Omega_m = 0.309 \pm 0.010, w_0 = -0.938 \pm 0.109, w_a = -0.38 \pm 0.41$.
\begin{equation}
 \Omega_m = 0.309 \pm 0.010, w_0 = -0.938 \pm 0.109, w_a = -0.38 \pm 0.41.
\end{equation}

To derive the constraints combined with our AP method,
we then compute the likelihood of the SDSS LOWZ+CMASS AP measurements for each of the parameter combinations provided in the Planck chain. 
To do this, we splited the 1,133,326 galaxies from the SDSS DR12 into six redshift bins,
measured the 2pCF (on scales of 6-40 $h^{-1}$Mpc) in these redshift bins (see Fig.\ref{fig_nbar}).
This enables us investigating the redshift evolution of anisotropic clustering.
We subtracted the systematic effects based on estimations from mocks made from Horizon Run 4 simulations \citep{HR4},
and used a $\chi^2$ function to assess the redshift dependence.
(see Methods for further explanations).

After adding the log-likelihoods of the Planck team sample with ours, 
while also multiply the sample weights by our likelihoods, 
we derive the CMB+BAO+SNIa+$H_0$+AP constraints on CPL parameters, %$\Omega_m = 0.301 \pm 0.008, w_0 = -1.042 \pm 0.067, w_a = -0.07 \pm 0.29$.
\begin{equation}
\Omega_m = 0.301 \pm 0.008, w_0 = -1.042 \pm 0.067, w_a = -0.07 \pm 0.29.
\end{equation}
While the central values of parameters do not shift appreciably,
the error bars of $w_0$ and $w_a$ are reduced by 38.4\% and 29.7\%, respectively.% $$-40\%$ by adding our method into the analysis.

@@

The presented method of clustering shells can be applied to many spectroscopic galaxy surveys. 
The method is complimentary to the BAO method and the two methods can be combined together, 
increasing (by roughly twice) the power of ongoing and future surveys, with little to no additional cost.
We expect this approach will play an important role in helping us unveiling the mystery of dark energy.

@@

\bibitem{CPL_CP}
Chevallier, M, \& Polarski, D. 2001, Int. J. Mod. Phys. D,  10, 213

\bibitem{CPL_L}
Linder, E. V. 2003, Phys. Rev. Lett, 90, 091301

@@

X.L. computed the approximated 2pCFs in the many cosmologies, did the statistical analysis and computed the likelihoods, 
derived the final cosmological constraints, and conducted the robustness tests. 
C.S. wrote the code for computing the 2pCFs, and contributed in the likelihood methodology and the robustness tests. 
C.P. contributed in the concept of using the redshift dependence to do AP test.
He also contributed in the statistical methods of LSS, the likelihood methodology,
the horizon run simulations, the systematic corrections, and the robustness tests.
Y.W. and G.Z. contributed in the computing of approximated 2pCFs in non-fiducial cosmologies and the robustness tests.
H.P. contributed in the 2pCF measurements, the likelihood computation, and the systematic correction.
A.S. contributed in the fast computation of likelihoods in non-fiducial cosmologies, and the cosmological interpretation of the results.
J.K. and S.H. performed the Horizon Run 4 simulations and created the mock galaxy samples.
All authors commented on the manuscript.

@@

In detail, the number of pairs are only counted in a fiducial cosmology
-- which is taken as the $\Omega_m=0.26$ $\Lambda$CDM --
and ``translated'' to the measurements in a ``target'' cosmology using the relations, 
\begin{eqnarray}
 s_{\rm target} = s_{\rm fiducial} \sqrt{\alpha_{\parallel}^2 \mu_{\rm fiducial}+\alpha_{\bot}(1-\mu_{\rm fiducial}^2)}, \\
 \mu_{\rm target} = \mu_{\rm fiducial} \frac{\alpha_\parallel}
 {\sqrt{\alpha_{\parallel}^2 \mu_{\rm fiducial} +\alpha_{\bot}(1-\mu_{\rm fiducial}^2)}}
\end{eqnarray}
where $\alpha_{\bot}\equiv D_{A,\rm target}/D_{A,\rm fiducial}$,
$\alpha_{\parallel}\equiv H_{\rm fiducial}/H_{\rm target}$.
and values of $D_A$ and $H$ are computed in the effective redshifts of the six redshift bins.
In the fiducial cosmology
we measure $\xi(s,\mu)$ in an ultra high resolution of
$\Delta s = 0.2 {\rm Mpc/h}$, $\Delta \mu = 1/600$,
%Let us call this ``approximate 2pCF'' method.
%to obtain the accurate number counts 
and these small ``pixels'' are grouped to infer 
the number counts in other cosmologies, 
in large pixels of $\Delta s = 1 {\rm Mpc/h}$, $\Delta \mu = 1/120$.
%in the fiducial cosmology and group the bins in the target cosmology.
In the case when one small pixel belongs to more than one larger pixel,
a correction is applied by computing the fraction of the overlapping area.

We found when using the approximate method the computed 
$\hat\xi_{\Delta s}(\mu)$ in a non-fiducial cosmology suffers from
an error of $\approx0.5\%$, which is 10 times smaller than 
the statistical noise in the 2pCF.
As a result, the approximate method reproduces the results presented in \cite{Li2016} quite well.
